---
title: "NYCFlights2"
output: html_document
---

### Work on data nycflights

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE}
library(statsr)
library(dplyr)
library(ggplot2)
```

### Departure delays over months

Which month would you expect to have the highest average delay departing 
from an NYC airport?

```{r mean-dep-delay-months}
nycflights %>%
  group_by(month) %>%
  summarise(mean_dd = mean(dep_delay)) %>%
  arrange(desc(mean_dd))
```

We can also visualize the distributions of departure delays across months using 
side-by-side box plots:

```{r delay-month-box}
ggplot(nycflights, aes(x = factor(month), y = dep_delay)) +
  geom_boxplot()
```

There is some new syntax here: We want departure delays on the y-axis and the
months on the x-axis to produce side-by-side box plots. Side-by-side box plots
require a categorical variable on the x-axis, however in the data frame `month` is 
stored as a numerical variable (numbers 1 - 12). Therefore we can force R to treat
this variable as categorical, what R calls a **factor**, variable with 
`factor(month)`.

### On time departure rate for NYC airports

Suppose you will be flying out of NYC and want to know which of the 
three major NYC airports has the best on time departure rate of departing flights. 
Suppose also that for you a flight that is delayed for less than 5 minutes is 
basically "on time". You consider any flight delayed for 5 minutes of more to be 
"delayed".

```{r dep-type}
nycflights <- nycflights %>%
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed"))
```

```{r}
nycflights %>%
  group_by(origin) %>%
  summarise(ot_dep_rate = sum(dep_type == "on time") / n()) %>%
  arrange(desc(ot_dep_rate))
```

We can also visualize the distribution of on on time departure rate across 
the three airports using a segmented bar plot.

```{r}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar()
```

 <!-- Add a column avg_speed -->
```{r mutate}
nycflights <- nycflights %>% 
  mutate(avg_speed=distance*60/air_time)
```

### View the maximum average speed
```{r max-avg_speed}
nycflights %>% 
  mutate(avg_speed=60*distance/air_time) %>% 
  select(avg_speed,tailnum) %>% 
  arrange(desc(avg_speed))

```


# Make a scatterplot of avg_speed vs. distance
```{r scatterplot}
ggplot(nycflights,aes(x=distance, y=avg_speed))+geom_point()
```

### What fraction of flights that were "delayed" departing arrive "on time"? 
```{r fraction}
nycflights <- nycflights %>%
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed"))%>% 
  mutate(arr_type=ifelse(arr_delay<=0,"on time","delayed"))
